A general and efficient representation of ancestral recombination graphs

### 2024-06-26 paper.tex
Update node labels
And change some captions to avoid colour mentions

@@

\newcommand{\noderef}[1]{\textsf{#1}}
\newcommand{\tsinfer}[0]{\texttt{tsinfer}}
\newcommand{\kwarg}[0]{\texttt{KwARG}}
\newcommand{\argweaver}[0]{\texttt{ARGweaver}}
\newcommand{\argweaverD}[0]{\texttt{ARGweaver-D}}
\newcommand{\relate}[0]{\texttt{Relate}}
\newcommand{\espalier}[0]{\texttt{Espalier}}
\newcommand{\arbores}[0]{\texttt{Arbores}}

@@

% \linenumbers
\section{Introduction}
\label{sec-intro}
% Para 1: ARGs are cool!
% NOTE: slight inaccuracy of "DNA" here offset by making the problem concrete.
Estimating the genetic genealogy of a set of DNA sequences
under the influence of recombination,
usually known as an Ancestral Recombination Graph (ARG), is a long-standing
goal in genetics.
Broadly speaking, an ARG describes the different paths of genetic inheritance
caused by recombination, encapsulating the resulting complex web of genetic
ancestors of a set of samples.
Recent breakthroughs
in large-scale inference
methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,
schaefer2021ancestral,wohns2022unified,zhang2023biobank,zhan2023towards,deng2024robust}
have raised the realistic prospect of ARG-based analysis becoming a standard part
of the population and statistical genetics toolkit~\citep{hejase2020summary}.
Applications using inferred ARGs as input have begun to
appear~\citep{osmond2021estimating,
fan2022genealogical,
hejase2022deep,
guo2022recombination,
zhang2023biobank,
nowbandegani2023extremely,
ignatieva2023distribution,
fan2023likelihood,
link2023tree,
grundler2024geographic,
huang2024estimating,
korfmann2024simultaneous,
deraje2024inferring}
and many more are sure to
follow~\citep{harris2019database,harris2023using}.

@@

\begin{figure*}
\begin{center}
    \includegraphics[width=\textwidth]{illustrations/arg-in-pedigree}
\end{center}
\caption{\label{fig-arg-in-pedigree}
An example genome ARG (gARG) embedded in a pedigree.
(A) Diploid individuals (blue), visualised in a highly inbred pedigree and
labelled $D_1$ to $D_8$,
contain both paternal and maternal  genomes
labelled \noderef{a} to \noderef{p}. Black lines show inheritance paths connecting
genomes in the current generation (\noderef{a} to \noderef{d}) with their ancestors.
Genomes \noderef{a} and \noderef{c} are the product of two independent
meioses (recombination events, in red) between
the paternal genomes \noderef{e}
and \noderef{f}, and regions of genome inherited are shown with shaded colour.
Genomes are shaded such that where, backwards in time,
they merge into a common ancestor, the merged region is darker.
(B) The corresponding gARG along with inheritance annotations on all edges
(partial inheritance in bold).
(C) The corresponding local trees.
}
\end{figure*}

As illustrated in Fig.~\ref{fig-arg-in-pedigree},
the gARG for a given set of individuals is embedded in their pedigree.
The figure shows the pedigree of eight diploid individuals
and their sixteen constituent genomes (each consisting of a single chromosome),
along with paths of genetic inheritance.
Here, and throughout,
nodes are labelled with lowercase alphabetical letters
rather than integer identifiers to avoid confusion with genomic intervals.
Thus individual $D_1$ is composed
of genomes \noderef{a} and \noderef{b}, which are inherited from its
two parents $D_3$ and $D_4$. Each inherited genome may be the recombined product
of the two genomes belonging to an individual parent.
In this example,
genome \noderef{b} was inherited directly from $D_4$'s genome \noderef{g} without
recombination, whereas
genome \noderef{a} is the recombinant product of
$D_2$'s genomes \noderef{e} and \noderef{f} crossing over at position 2.
Specifically, genome \noderef{a} inherited the (half-closed)
interval $[0, 2)$ from genome \noderef{e} and $[2, 10)$ from genome \noderef{f}.
These intervals are shown attached to the corresponding graph edges.
The figure shows the annotated pedigree with realised inheritance of genomes
between generations (A), the corresponding gARG (B), and finally the corresponding
sequence of local trees along the
genome (C).
The local trees span the three genome regions delineated
by the two recombination breakpoints that gave rise to these genomes;
see Appendix~\ref{sec-ARG-and-local-trees} for details
on how local trees are embedded in an ARG.

@@

\section{Event ARGs}
\label{sec-eARG}
A classical view of an ARG data structure,
described explicitly in several publications~\citep[e.g.][]{
wiuf1999recombination,gusfield2014recombinatorics,hayman2023recoverability},
interprets nodes not as genomes but as historical \emph{events}
(but see~\citet{parida2011minimal} and \citet{zhang2023biobank}
for notable exceptions).
This Event ARG (eARG) encoding is the basis of the output formats 
created by multiple ARG inference tools
\citep[e.g.][]{song2004minimum,song2005efficient,rasmussen2014genome,
heine2018bridging,ignatieva2021kwarg}.
% Griffiths never mentions "sampling events", so let's not get
% into that. It's not an important detail, and we don't want to be
% overly precise about something that people haven't actually written
% down.
In this encoding there are two types of internal node in the graph,
representing the most recent common ancestor and recombination events
in the history of a sample.
At common ancestor nodes, the inbound lineages merge into a
single ancestral lineage with one parent, and at recombination
nodes a single lineage is split into two independent
ancestral lineages. Recombination nodes are annotated with
the corresponding crossover breakpoints, and these breakpoints
are used to construct the local trees.
This is done by tracing pastwards through the graph from the samples,
making decisions about which outbound edge to follow through
recombination nodes based on the breakpoint
position~\citep{griffiths1996ancestral}. Fig.~\ref{fig-event-arg} shows an example of an
eARG with three sample genomes (\noderef{a}, \noderef{b}, and \noderef{c}),
three common ancestor events (\noderef{e}, \noderef{f}, and \noderef{g})
and a single recombination event (node \noderef{d}) with a breakpoint
at position $x$.
Assigning a breakpoint to a recombination node is
not sufficient to uniquely define the local trees, and either
some additional ordering rules~\citep[e.g.][]{griffiths1996ancestral} or
explicit information~\citep[e.g.][]{gusfield2014recombinatorics,ignatieva2021kwarg}
is required to distinguish the left and right parents.
We assume in Fig.~\ref{fig-event-arg} that \noderef{d} inherits genetic material to the
left of $x$ from \noderef{e} and to the right of $x$ from \noderef{f}.

\begin{figure}
\centering
\tikzmath{\x1 =0; \x2=8;\xx=12.5; \x3=14; \xt=18;}
\begin{tikzpicture}[x=5mm, y=5mm, node distance=2mm and 20mm]
\tikzset{greynode/.style={circle,fill,inner sep=1},
nodelabel/.style={font=\footnotesize}}

\node [anchor=north west] at (\x1,6) {A};
\node [anchor=north west] at (\x2,6) {B};
% \node [anchor=north west] at (\xt,6) {C};

%%% (A) ARG

@@

%%% (B) Trees
\node (l0) [greynode] at (\x2 + 0, 0) {};
\node (l1) [greynode] at (\x2 + 2, 0) {};
\node (l2) [greynode] at (\x2 + 3, 0) {};
\node (l3) [greynode] at (\x2 + 1, 2) {};
\node (l4) [greynode] at (\x2 + 2, 4) {};

@@

\foreach \u/\lab in {
        s0/$\noderef{a}$, s1/$\noderef{b}$, s2/$\noderef{c}$,
        r0/$\noderef{a}$, r1/$\noderef{b}$, r2/$\noderef{c}$,
        l0/$\noderef{a}$, l1/$\noderef{b}$, l2/$\noderef{c}$}
    \node[nodelabel,anchor=south] at ([yshift=-12pt]\u) {\lab};

\foreach \u/\lab in {
        s6/$\noderef{g}$,
        l4/$\noderef{g}$,
        r4/$\noderef{g}$}
    \node[nodelabel,anchor=south] at (\u) {\lab};

\node [nodelabel,anchor=north west] at ($(s3) + (\x1 + 0,0)$) {$x$};
\foreach \u/\lab in {
        s4/$\noderef{e}$,
        l3/$\noderef{e}$}
    \node[nodelabel,anchor=south west] at (\u) {\lab};
\foreach \u/\lab in {
        s5/$\noderef{f}$,
        r3/$\noderef{f}$}
    \node[nodelabel,anchor=south east] at (\u) {\lab};
\foreach \u/\lab in {s3/$\noderef{d}$, s6/$\noderef{g}$}
    \node[nodelabel,anchor=south] at (\u) {\lab};

@@

% %%% (C) Encoding
% \node [nodelabel,anchor=north west] at ($(\xt,5)$) {
% \begin{tabular}{c|c|l}
% % \multicolumn{2}{c}{Breakpoints}\\
% Node & Breakpoint & Parents\\
% \hline
% $\noderef{a}$ & $\varnothing$ & [\noderef{e}]\\
% $\noderef{b}$ & $\varnothing$ & [\noderef{d}] \\
% $\noderef{c}$ & $\varnothing$ & [\noderef{f}]\\
% $\noderef{d}$ & $x$ & [\noderef{e}, \noderef{f}] \\
% $\noderef{e}$ & $\varnothing$ & [\noderef{g}]\\
% $\noderef{f}$ & $\varnothing$ & [\noderef{g}]\\
% $\noderef{g}$ & $\varnothing$ & []\\
% \end{tabular}};

\end{tikzpicture}
\caption{\label{fig-event-arg}
A classical event ARG (eARG). (A) Standard graph depiction with
breakpoint $x$ associated with the recombination node \noderef{d}.
Nodes \noderef{e}, \noderef{f} and \noderef{g} are common ancestor events.
(B) Corresponding local trees to the left and right of breakpoint $x$
(note these are shown in the conventional form in which only coalescences
within the local tree are included; see Appendix~\ref{sec-ARG-and-local-trees}
for a discussion of this important point).
}
\end{figure}

@@

\section{Ancestral material and sample resolution}
\label{sec-ancestral-material}
Ancestral material~\citep{wiuf1999ancestry,wiuf1999recombination}
is a key concept in understanding the overall inheritance structure
of an ARG.
It denotes the genomic intervals ancestral to a set of samples
on the edges of an ARG.
For example, in Fig.~\ref{fig-arg-in-pedigree} we have
four sample genomes, \noderef{a}--\noderef{d}. As we
trace their genetic ancestry into the previous generation
(\noderef{e}--\noderef{h}), we can think of their ancestral
material propagating through the graph
pastwards. In the region $[2, 7)$, there is a
local coalescence where nodes \noderef{a} and \noderef{c}
find a common ancestor in \noderef{f}. Thus, in this region,
we have three genome segments that are ancestral to the
four samples. Fig.~\ref{fig-arg-in-pedigree}A
illustrates this by (shaded) ancestral material being present
in only three nodes (\noderef{f}, \noderef{g}, and \noderef{h}) in this region,
while node \noderef{e} is blank
as it carries \emph{non-ancestral} material.
This process of local coalescence continues through the
graph, until all samples reach their most recent common
ancestor in node \noderef{n}.

@@

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{illustrations/ancestry-resolution}
\caption{\label{fig-ancestry-resolution}
Converting the \citet[][Fig.~1]{wiuf1999recombination} example
to a sample-resolved gARG. (A) The original eARG; square nodes
represent sampling (black), common ancestor (blue), and recombination (red) events;
the latter contain breakpoint positions.
(B) The corresponding gARG with breakpoints directly converted to
edges annotated with inheritance intervals.
(C) The sample-resolved gARG resulting from simplifying with respect
to the sample genomes, \noderef{a}, \noderef{b}, and \noderef{c}.
Dashed lines show edges that are
no longer present (in practice, nodes \noderef{g}, \noderef{j}, and \noderef{q} would also be removed).
Coalescence with respect to the sample is indicated by shaded bars, as
in Fig.~\ref{fig-arg-in-pedigree}A; nodes \noderef{n}, \noderef{o}, \noderef{p}, \noderef{q} have truncated
bars showing that local ancestry of entirely coalesced regions is omitted.
Line thickness is proportional to the genomic span of each edge.
Nodes representing recombination events are retained
for clarity, but could be removed by simplification if
desired.
}
\end{figure*}

Any eARG can be converted to a sample-resolved gARG
via a two-step process illustrated in Fig.~\ref{fig-ancestry-resolution}.
The first step is to take the input eARG (Fig.~\ref{fig-ancestry-resolution}A),
duplicate its graph topology, and then add inheritance annotations
to each of the gARG's edges (Fig.~\ref{fig-ancestry-resolution}B) as follows.
If a given node is a common ancestor event, we annotate the single
outbound edge with the interval $[0,L)$, for a genome of length $L$. If the
node is a recombination event with a breakpoint $x$, we annotate the two
outbound edges respectively with the intervals $[0, x)$ and $[x, L)$. These
inheritance interval annotations are clearly in one-to-one correspondence with
the information in the input eARG. They are also analogous to the
inheritance intervals we get on the edges in a prospective gARG
produced by a forward-time simulation, which are concerned with recording
the direct genetic relationship between a parent and child genome and are not
necessarily minimal in terms of the ancestral material of a sample.
Thus, the final step is to use the ``simplify'' algorithm to resolve the
ancestral material of the samples (Fig.~\ref{fig-ancestry-resolution}C).

The sample-resolved gARG of Fig.~\ref{fig-ancestry-resolution}C
differs in some important ways to the
original eARG (Fig.~\ref{fig-ancestry-resolution}A).
Firstly, we can see that some nodes and edges have been removed entirely
from the graph.
The ``grand MRCA'' \noderef{q} is omitted from the
sample-resolved gARG because all segments of the genome have
fully coalesced in \noderef{k} and \noderef{p} before \noderef{q} is reached.
Likewise, the edge
between \noderef{g} and \noderef{j} is omitted because the recombination
event at position $5$ (represented by node \noderef{g})
fell in non-ancestral material.
More generally, we can see that the sample resolved
gARG of Fig.~\ref{fig-ancestry-resolution}C
allows for ``local'' inspection
of an ARG in a way that is not possible in an eARG.
Because the ancestral material is stored with each edge of a gARG, the
cumulative effects of events over time can be reasoned
about, without first ``replaying'' those events.
Many computations
that we wish to perform on an ARG will require resolving
the ancestral material with respect to a set of samples.
% this is weak, but would be good to emphasise this point.
% you just end up running simplify lots of times, in your
% downstream algs
The gARG encoding
allows us to perform this once
and to store the result,
whereas the eARG encoding requires us to repeat the process
each time.

@@

\section{A diversity of structures}
\begin{figure*} \begin{center}
\includegraphics[width=\textwidth]{illustrations/inference.pdf} \end{center}
\caption{\label{fig-inferred-args} Inference of sample-resolved ARGs for 11
\textit{Drosophila melanogaster} DNA sequences over a 2.4kb
region of the ADH locus~\citep{kreitman1983nucleotide}.
Results for four different methods:
(A) \kwarg; (B) \argweaver; (C) \tsinfer; and (D) \relate, converted 
to the standard \emph{tskit} gARG encoding. See
Appendix \ref{sec-example-inferred-args} for details of these methods.
Edge colours indicate time of the edge's child node
(lighter: older; darker: younger).
Vertical and horizontal positions of graph nodes are arbitrary.
Line width and node colour are as described in Fig.~\ref{fig-simplification}.
Bottom row graphics show the genome positions, relative to the start of
the ADH gene, for each graph edge from the corresponding ARG. Edge intervals
are drawn as horizontal lines, stacked in time order (edges with youngest
children at the bottom); vertical dashed lines denote breakpoints between
local trees.
}
\end{figure*}
A key goal of this perspective is to highlight the heterogeneity of
the graph structures inferred by modern ARG inference methods.
To illustrate this point, Fig.~\ref{fig-inferred-args} shows the 
output of 
\kwarg~\citep{ignatieva2021kwarg},
\argweaver~\citep{rasmussen2014genome},
\tsinfer~\citep{kelleher2019inferring},
and \relate~\citep{speidel2019method}
on the classical \cite{kreitman1983nucleotide} dataset. 
The ARGs in Fig.~\ref{fig-inferred-args}A and B are precise 
estimates (Appendix~\ref{sec-precision}),
with each node corresponding to a common ancestor 
or recombination event, or  equivalently, either having two children
or two parents.
In contrast the ARGs in Fig.~\ref{fig-inferred-args}C and D do not 
have this clear-cut interpretation, and the nodes
can simultaneously have more that than two children and more
than two parents. Another dimension of variability among the ARGs
is that the first three methods infer nodes that have
a ``coalescence span''  greater than 0 and less than 100\%, indicating that there 
are nodes that are ``locally unary'' (Appendix~\ref{sec-locally-unary-edges}),
but mark a coalescence between lineages elsewhere along the sequence.

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{illustrations/simplification}
\caption{\label{fig-simplification}
Levels of ARG simplification.
(A) An example gARG simulated from a diploid Wright-Fisher model.
(B) Remove all
singly-connected graph components (e.g., diamonds such as \noderef{jlnm}).
(C) Remove nodes that never represent coalescences,
i.e.\ are unary everywhere (e.g.\ \noderef{n} and \noderef{r}).
(D) Rewrite edges to bypass nodes in local trees in which they are unary
(often described as ``fully simplified'').
In each case, the graph is shown on the left
and corresponding local trees on the right.
% Repetition here, but I'm thinking about the casual "I'm just reading
% the captions reader" who might not realise that there's an essential
% ingredient missing from the graph version
In the interest of visual clarity, inheritance intervals are not shown
on the graph edges; Supplementary Fig.~S1
shows the graphs with these inheritance intervals included.
Graph nodes are coloured by the number of parents and shaded
according to the proportion of their span over which they are coalescent;
see the text for more details.
}
\end{figure*}

>>>>

% Uncommment to extract tikz pictures
% \usetikzlibrary{external}
% \tikzexternalize

@@

\newcommand{\noderef}[1]{\textit{#1}}
\newcommand{\tsinfer}[0]{\texttt{tsinfer}}
\newcommand{\kwarg}[0]{\texttt{KwARG}}
\newcommand{\argweaver}[0]{\texttt{ARGweaver}}
\newcommand{\argweaverD}[0]{\texttt{ARGweaver-D}}
\newcommand{\relate}[0]{\texttt{Relate}}
\newcommand{\espalier}[0]{\texttt{Espalier}}
\newcommand{\arbores}[0]{\texttt{Arbores}}

@@

% \linenumbers
\section{Introduction}
\label{sec-intro}
% Para 1: ARGs are cool!
% NOTE: slight inaccuracy of "DNA" here offset by making the problem concrete.
Estimating the genetic genealogy of a set of DNA sequences
under the influence of recombination,
usually known as an Ancestral Recombination Graph (ARG), is a long-standing
goal in genetics.
Broadly speaking, an ARG describes the different paths of genetic inheritance
caused by recombination, encapsulating the resulting complex web of genetic
ancestors of a set of sampled genomes.
Recent breakthroughs
in large-scale inference
methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,
schaefer2021ancestral,wohns2022unified,zhang2023biobank,zhan2023towards,deng2024robust}
have raised the realistic prospect of ARG-based analysis becoming a standard part
of the population and statistical genetics toolkit~\citep{hejase2020summary}.
Applications using inferred ARGs as input have begun to
appear~\citep{osmond2021estimating,
fan2022genealogical,
hejase2022deep,
guo2022recombination,
zhang2023biobank,
nowbandegani2023extremely,
ignatieva2023distribution,
fan2023likelihood,
link2023tree,
grundler2024geographic,
huang2024estimating,
korfmann2024simultaneous,
deraje2024inferring}
and many more are sure to
follow~\citep{harris2019database,harris2023using}.

@@

\begin{figure*}
\begin{center}
    \includegraphics[width=\textwidth]{illustrations/arg-in-pedigree}
\end{center}
\caption{\label{fig-arg-in-pedigree}
An example genome ARG (gARG) embedded in a pedigree.
\textbf{a.} Diploid individuals (shaded background squares / circles), visualised in a highly inbred pedigree and
labelled $d_1$ to $d_8$,
contain both paternal and maternal  genomes
labelled \noderef{A} to \noderef{P}. Lines show inheritance paths connecting
genomes in the current generation (\noderef{A} to \noderef{D}) with their ancestors.
Genomes \noderef{A} and \noderef{C} are the product of two independent
meioses (recombination events, with italicised breakpoint positions) between
the paternal genomes \noderef{E}
and \noderef{F}; the regions of inherited genome are shaded.
Further shading highlights genomic regions that merge into a common ancestor:
merged regions gradually darken until fully coalesced.
\textbf{b.} The corresponding gARG along with inheritance annotations on all edges
(partial genomic inheritance in bold).
\textbf{c.} The corresponding local trees.
}
\end{figure*}

As illustrated in Fig.~\ref{fig-arg-in-pedigree},
the gARG for a given set of individuals is embedded in their pedigree.
The figure shows the pedigree of eight diploid individuals
and their sixteen constituent genomes (each consisting of a single chromosome),
along with paths of genetic inheritance.
Here, and throughout,
nodes are labelled with uppercase alphabetical letters
rather than integer identifiers to avoid confusion with genomic intervals.
Thus individual $d_1$ is composed
of genomes \noderef{A} and \noderef{B}, which are inherited from its
two parents $d_3$ and $d_4$. Each inherited genome may be the recombined product
of the two genomes belonging to an individual parent.
In this example,
genome \noderef{B} was inherited directly from $d_4$'s genome \noderef{G} without
recombination, whereas
genome \noderef{A} is the recombinant product of
$d_2$'s genomes \noderef{E} and \noderef{F} crossing over at position 2.
Specifically, genome \noderef{A} inherited the (half-closed)
interval $[0, 2)$ from genome \noderef{E} and $[2, 10)$ from genome \noderef{F}.
These intervals are shown attached to the corresponding graph edges.
The figure shows the annotated pedigree with realised inheritance of genomes
between generations (a), the corresponding gARG (b), and finally the corresponding
sequence of local trees along the
genome (c).
The local trees span the three genome regions delineated
by the two recombination breakpoints that gave rise to these genomes;
see Appendix~\ref{sec-ARG-and-local-trees} for details
on how local trees are embedded in an ARG.

@@

\section{Event ARGs}
\label{sec-eARG}
A classical view of an ARG data structure,
described explicitly in several publications~\citep[e.g.][]{
wiuf1999recombination,gusfield2014recombinatorics,hayman2023recoverability},
interprets nodes not as genomes but as historical \emph{events}
(but see~\citet{parida2011minimal} and \citet{zhang2023biobank}
for notable exceptions).
This Event ARG (eARG) encoding is the basis of the output formats 
created by multiple ARG inference tools
\citep[e.g.][]{song2004minimum,song2005efficient,rasmussen2014genome,
heine2018bridging,ignatieva2021kwarg}.
% Griffiths never mentions "sampling events", so let's not get
% into that. It's not an important detail, and we don't want to be
% overly precise about something that people haven't actually written
% down.
In this encoding there are two types of internal node in the graph,
representing the most recent common ancestor and recombination events
in the history of a sample.
At common ancestor nodes, the inbound lineages merge into a
single ancestral lineage with one parent, and at recombination
nodes a single lineage is split into two independent
ancestral lineages. Recombination nodes are annotated with
the corresponding crossover breakpoints, and these breakpoints
are used to construct the local trees.
This is done by tracing pastwards through the graph from the samples,
making decisions about which outbound edge to follow through
recombination nodes based on the breakpoint
position~\citep{griffiths1996ancestral}. Fig.~\ref{fig-event-arg} shows an example of an
eARG with three sample genomes (\noderef{A}, \noderef{B}, and \noderef{C}),
three common ancestor events (\noderef{E}, \noderef{F}, and \noderef{G})
and a single recombination event (node \noderef{D}) with a breakpoint
at position $x$.
Assigning a breakpoint to a recombination node is
not sufficient to uniquely define the local trees, and either
some additional ordering rules~\citep[e.g.][]{griffiths1996ancestral} or
explicit information~\citep[e.g.][]{gusfield2014recombinatorics,ignatieva2021kwarg}
is required to distinguish the left and right parents.
We assume in Fig.~\ref{fig-event-arg} that \noderef{D} inherits genetic material to the
left of $x$ from \noderef{E} and to the right of $x$ from \noderef{F}.

\begin{figure}
\centering
\tikzmath{\x1 =0; \x2=8.5;\xx=12.5; \x3=13.5; \xt=18;}
\begin{tikzpicture}[x=5mm, y=5mm, node distance=2mm and 20mm]
\tikzset{greynode/.style={circle,fill,inner sep=1},
nodelabel/.style={font=\footnotesize}}

\node [anchor=north west] at (\x1-.5,5.5) {\textsf{\textbf{\LARGE a}}};
\node [anchor=north west] at (\x2-.5,5.5) {\textsf{\textbf{\LARGE b}}};
% \node [anchor=north west] at (\xt,6) {C};

%%% (a) ARG

@@

%%% (b) Trees
\node (l0) [greynode] at (\x2 + 0, 0) {};
\node (l1) [greynode] at (\x2 + 2, 0) {};
\node (l2) [greynode] at (\x2 + 3, 0) {};
\node (l3) [greynode] at (\x2 + 1, 2) {};
\node (l4) [greynode] at (\x2 + 2, 4) {};

@@

\foreach \u/\lab in {
        s0/$\noderef{A}$, s1/$\noderef{B}$, s2/$\noderef{B}$,
        r0/$\noderef{A}$, r1/$\noderef{B}$, r2/$\noderef{B}$,
        l0/$\noderef{A}$, l1/$\noderef{B}$, l2/$\noderef{B}$}
    \node[nodelabel,anchor=south] at ([yshift=-12pt]\u) {\lab};

\foreach \u/\lab in {
        s6/$\noderef{G}$,
        l4/$\noderef{G}$,
        r4/$\noderef{G}$}
    \node[nodelabel,anchor=south] at (\u) {\lab};

\node [nodelabel,anchor=north west] at ($(s3) + (\x1 + 0,0)$) {$x$};
\foreach \u/\lab in {
        s4/$\noderef{E}$,
        l3/$\noderef{E}$}
    \node[nodelabel,anchor=south west] at (\u) {\lab};
\foreach \u/\lab in {
        s5/$\noderef{F}$,
        r3/$\noderef{F}$}
    \node[nodelabel,anchor=south east] at (\u) {\lab};
\foreach \u/\lab in {s3/$\noderef{D}$, s6/$\noderef{G}$}
    \node[nodelabel,anchor=south] at (\u) {\lab};

@@

% %%% (c) Encoding
% \node [nodelabel,anchor=north west] at ($(\xt,5)$) {
% \begin{tabular}{c|c|l}
% % \multicolumn{2}{c}{Breakpoints}\\
% Node & Breakpoint & Parents\\
% \hline
% $\noderef{A}$ & $\varnothing$ & [\noderef{E}]\\
% $\noderef{B}$ & $\varnothing$ & [\noderef{D}] \\
% $\noderef{C}$ & $\varnothing$ & [\noderef{F}]\\
% $\noderef{D}$ & $x$ & [\noderef{E}, \noderef{F}] \\
% $\noderef{E}$ & $\varnothing$ & [\noderef{G}]\\
% $\noderef{F}$ & $\varnothing$ & [\noderef{G}]\\
% $\noderef{G}$ & $\varnothing$ & []\\
% \end{tabular}};

\end{tikzpicture}
\caption{\label{fig-event-arg}
A classical event ARG (eARG). \textbf{a.} Standard graph depiction with
breakpoint $x$ associated with the recombination node \noderef{D}.
Nodes \noderef{E}, \noderef{F} and \noderef{G} are common ancestor events.
\textbf{b.} Corresponding local trees to the left and right of breakpoint $x$
(note these are shown in the conventional form in which only coalescences
within the local tree are included, hence \noderef{D} is omitted; see
Appendix~\ref{sec-ARG-and-local-trees} for a discussion of this important point).
}
\end{figure}

@@

\section{Ancestral material and sample resolution}
\label{sec-ancestral-material}
Ancestral material~\citep{wiuf1999ancestry,wiuf1999recombination}
is a key concept in understanding the overall inheritance structure
of an ARG.
It denotes the genomic intervals ancestral to a set of samples
on the edges of an ARG.
For example, in Fig.~\ref{fig-arg-in-pedigree} we have
four sample genomes, \noderef{A}--\noderef{D}. As we
trace their genetic ancestry into the previous generation
(\noderef{E}--\noderef{H}), we can think of their ancestral
material propagating through the graph
pastwards. In the region $[2, 7)$, there is a
local coalescence where nodes \noderef{A} and \noderef{C}
find a common ancestor in \noderef{F}. Thus, in this region,
we have three genome segments that are ancestral to the
four samples. Fig.~\ref{fig-arg-in-pedigree}a
illustrates this by (shaded) ancestral material being present
in only three nodes (\noderef{F}, \noderef{G}, and \noderef{H}) in this region,
while node \noderef{E} is blank
as it carries \emph{non-ancestral} material.
This process of local coalescence continues through the
graph, until all samples reach their most recent common
ancestor in node \noderef{N}.

@@

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{illustrations/ancestry-resolution}
\caption{\label{fig-ancestry-resolution}
Converting the \citet[][Fig.~1]{wiuf1999recombination} example
to a sample-resolved gARG. \textbf{a.} The original eARG; nodes
represent sampling, common ancestor, and recombination events (small shaded, small blue, and large red rectangles respectively); the latter contain breakpoint positions.
\textbf{b.} The corresponding gARG with breakpoints directly converted to
edges annotated with inheritance intervals.
\textbf{c.} The sample-resolved gARG resulting from simplifying with respect
to the sample genomes, \noderef{A}, \noderef{B}, and \noderef{C}.
Dashed lines show edges that are
no longer present (in practice, nodes \noderef{G}, \noderef{J}, and \noderef{Q} would also be removed).
Coalescence with respect to the sample is indicated by shaded bars, as
in Fig.~\ref{fig-arg-in-pedigree}a; nodes \noderef{N}, \noderef{O}, \noderef{P}, \noderef{Q} have truncated
bars showing that local ancestry of entirely coalesced regions is omitted.
Line thickness is proportional to the genomic span of each edge.
Nodes representing recombination events are retained
for clarity, but could be removed by simplification if
desired.
}
\end{figure*}

Any eARG can be converted to a sample-resolved gARG
via a two-step process illustrated in Fig.~\ref{fig-ancestry-resolution}.
The first step is to take the input eARG (Fig.~\ref{fig-ancestry-resolution}a),
duplicate its graph topology, and then add inheritance annotations
to each of the gARG's edges (Fig.~\ref{fig-ancestry-resolution}b) as follows.
If a given node is a common ancestor event, we annotate the single
outbound edge with the interval $[0,L)$, for a genome of length $L$. If the
node is a recombination event with a breakpoint $x$, we annotate the two
outbound edges respectively with the intervals $[0, x)$ and $[x, L)$. These
inheritance interval annotations are clearly in one-to-one correspondence with
the information in the input eARG. They are also analogous to the
inheritance intervals we get on the edges in a prospective gARG
produced by a forward-time simulation, which are concerned with recording
the direct genetic relationship between a parent and child genome and are not
necessarily minimal in terms of the ancestral material of a sample.
Thus, the final step is to use the ``simplify'' algorithm to resolve the
ancestral material of the samples (Fig.~\ref{fig-ancestry-resolution}c).

The sample-resolved gARG of Fig.~\ref{fig-ancestry-resolution}c
differs in some important ways to the
original eARG (Fig.~\ref{fig-ancestry-resolution}a).
Firstly, we can see that some nodes and edges have been removed entirely
from the graph.
The ``grand MRCA'' \noderef{Q} is omitted from the
sample-resolved gARG because all segments of the genome have
fully coalesced in \noderef{K} and \noderef{P} before \noderef{Q} is reached.
Likewise, the edge
between \noderef{G} and \noderef{J} is omitted because the recombination
event at position $5$ (represented by node \noderef{G})
fell in non-ancestral material.
More generally, we can see that the sample resolved
gARG of Fig.~\ref{fig-ancestry-resolution}c
allows for ``local'' inspection
of an ARG in a way that is not possible in an eARG.
Because the ancestral material is stored with each edge of a gARG, the
cumulative effects of events over time can be reasoned
about, without first ``replaying'' those events.
Many computations
that we wish to perform on an ARG will require resolving
the ancestral material with respect to a set of samples.
% this is weak, but would be good to emphasise this point.
% you just end up running simplify lots of times, in your
% downstream algs
The gARG encoding
allows us to perform this once
and to store the result,
whereas the eARG encoding requires us to repeat the process
each time.

@@

\section{A diversity of structures}
\begin{figure*} \begin{center}
\includegraphics[width=\textwidth]{illustrations/inference.pdf} \end{center}
\caption{\label{fig-inferred-args} Inference of sample-resolved ARGs for 11
\textit{Drosophila melanogaster} DNA sequences over a 2.4kb
region of the ADH locus~\citep{kreitman1983nucleotide}.
Results for four different methods:
\textbf{a.} \kwarg; \textbf{b.} \argweaver; \textbf{c.} \tsinfer; and \textbf{d.} \relate, converted 
to the standard \emph{tskit} gARG encoding. See
Appendix \ref{sec-example-inferred-args} for details of these methods.
Edge colours indicate time of the edge's child node
(lighter: older; darker: younger).
Vertical and horizontal positions of graph nodes are arbitrary.
Line width and node colour are as described in Fig.~\ref{fig-simplification}.
Bottom row graphics show the genome positions, relative to the start of
the ADH gene, for each graph edge from the corresponding ARG. Edge intervals
are drawn as horizontal lines, stacked in time order (edges with youngest
children at the bottom); vertical dashed lines denote breakpoints between
local trees.
}
\end{figure*}
A key goal of this perspective is to highlight the heterogeneity of
the graph structures inferred by modern ARG inference methods.
To illustrate this point, Fig.~\ref{fig-inferred-args} shows the 
output of 
\kwarg~\citep{ignatieva2021kwarg},
\argweaver~\citep{rasmussen2014genome},
\tsinfer~\citep{kelleher2019inferring},
and \relate~\citep{speidel2019method}
on the classical \cite{kreitman1983nucleotide} dataset. 
The ARGs in Fig.~\ref{fig-inferred-args}a, b are precise 
estimates (Appendix~\ref{sec-precision}),
with each node corresponding to a common ancestor 
or recombination event, or  equivalently, either having two children
or two parents.
In contrast the ARGs in Fig.~\ref{fig-inferred-args}c, d do not 
have this clear-cut interpretation, and the nodes
can simultaneously have more that than two children and more
than two parents. Another dimension of variability among the ARGs
is that the first three methods infer nodes that have
a ``coalescence span''  greater than 0 and less than 100\%, indicating that there 
are nodes that are ``locally unary'' (Appendix~\ref{sec-locally-unary-edges}),
but mark a coalescence between lineages elsewhere along the sequence.

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{illustrations/simplification}
\caption{\label{fig-simplification}
Levels of ARG simplification.
\textbf{a.} An example gARG simulated from a diploid Wright-Fisher model.
\textbf{b.} Remove all
singly-connected graph components (e.g., diamonds such as \noderef{JLMN}).
\textbf{c.} Remove nodes that never represent coalescences,
i.e.\ are unary everywhere (e.g.\ \noderef{N} and \noderef{R}).
\textbf{d.} Rewrite edges to bypass nodes in local trees in which they are unary
(often described as ``fully simplified'').
In each case, the graph is shown on the left
and corresponding local trees on the right.
% Repetition here, but I'm thinking about the casual "I'm just reading
% the captions reader" who might not realise that there's an essential
% ingredient missing from the graph version
In the interest of visual clarity, inheritance intervals are not shown
on the graph edges; Supplementary Fig.~S1
shows the graphs with these inheritance intervals included.
Graph nodes are coloured by the number of parents and shaded
according to the proportion of their span over which they are coalescent;
see the text for more details.
}
\end{figure*}
